Distributional Trees for Circular Data

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Update application and include caption (first draft)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2428 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3435cd96e290c5557cc826f003664d337ed4b705

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
\label{fig:tree}
\end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
presents an application of circular regression trees  on hourly direction
forecasts at Innsbruck Airport.  Innsbruck lies at the bottom of the Inn
Valley, a deep mountain valley in the Eastern European Alps orientated parallel
to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
in the North with a approx. 2300\,m high mountain ridge and opens to the South
into the crest-perpendicular Wipp Valley. As potential split variables we
employ several 1-hourly lagged observed weather variables (e.g., wind
direction, wind speed, relative humidity, pressure gradients along the valley)
and derived spatial and temporal difference of these lagged observations.

As an illustration of the tree based wind direction forecasts at Innsbruck
airport, a BIC based post-pruned regression tree of depth three is shown in
Figure~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
The final node 11 with a concentration parameter $\kappa$ of only $0.84$
captures the cases with rather low wind speeds which typically don't show a
clear wind direction or distinct wind regime. In general, winds typically blow
from areas of high pressure to areas of low pressure where strong pressure
gradients yield high wind speeds and a rather well-defined wind direction.
Thus, the splits in the lagged wind direction perform a clustering into the
different wind regimes and the additional splits in pressure gradients or wind
speed divide the predictions according to the expected consistency of the wind
direction. This can be seen in the final nodes with strong pressure gradients
or high wind speeds where most training cases are classified correctly and
therefore the concentration parameter as an indicator of the certainty of the
forecast is rather large.

>>>>

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}

\caption{Fitted distribution tree for a circular response employing the von
Mises distribution. For each terminal node, the empirical histogram and the
fitted density distribution is displayed in gray and red, respectively. The
location parameters are depicted as red hands and convey the direction from
which the wind is blowing. The chosen split variables are wind direction
(dd~[$\text{rad}$]), wind speed (ff~[$\text{ms}^{-1}$]), and pressure gradients
along the upper (pdiff\_up~[$\text{hPa}$]) and lower
(pdiff\_low[$\text{hPa}$]) Inn Valley where positive values point in the
direction of Innsbruck.}
\label{fig:tree} \end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
presents an application of circular regression trees  on hourly direction
forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
Valley, a deep mountain valley in the Eastern European Alps orientated parallel
to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
in the North with a approx. 2300\,m high mountain ridge and opens to the South
into the crest-perpendicular Wipp Valley. As potential split variables we
employ several 1-hourly lagged observed weather variables, amongst others, the
pressure gradient along the upper and lower Inn Valley, the wind direction, and
the wind speed at Innsbruck airport.

As an illustration of the tree based wind direction forecasts at Innsbruck
airport, a regression tree of depth three is depicted in Figure~\ref{fig:tree}
showing the empirical histogram and fitted response distribution for each terminal node.
According to the fitted location parameters the following wind regimes can be summarized:
Firstly, up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). Secondly, downslope winds blowing across
the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). Thirdly, down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). The terminal node
11 with a concentration parameter $\kappa$ of only $0.84$ captures the cases
with rather low wind speeds which typically do not show a clear wind direction
or distinct wind regime. In general, winds typically blow from areas of high
pressure to areas of low pressure where strong pressure gradients yield high
wind speeds and a rather well-defined wind direction. Thus, the splits in the
lagged wind direction perform a clustering into the different wind regimes
employed by the fitted location parameters, and the additional splits in
pressure gradients or wind speed divide the predictions according to the
expected consistency of the wind direction expressed by the fitted
concentration parameters. This can be seen in the terminal nodes with strong
pressure gradients or high wind speeds where most training cases are classified
correctly and therefore the concentration parameter as an indicator of the
certainty of the forecast is rather large.
